Linear instability underlying synchronous waving of marine grass in flow

### 2014-09-07 Paper/grass.tex
pull from shreyas

\documentclass[aps,prl,reprint,twocolumn,showpacs,superscriptaddress,10pt]{revtex4-1}  % for review and submission
%\documentclass[aps,twocolumn,floatfix,prl,10pt]{revtex4-1}
%\documentclass[letterpaper,10pt,prl,twocolumn,aps]{revtex4-1}
%\documentclass[aps,preprint,floatfix,prl]{revtex4-1}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color}
\usepackage{longtable}
\usepackage{array}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\graphicspath{ {images/}{images/Growth_Rate_Cont_Plots/} }
\newcommand{\bn}{{\boldsymbol{\hat{n}}}}
\newcommand{\bt}{{\boldsymbol{\hat{t}}}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand{\del}{\partial}
\newcommand{\hg}{h_g}
\newcommand{\Rey}{\text{R}}
\newcommand{\Ndg}{\tilde{N}_g}
\newcommand{\monami}{\textit{monami}}
% \newcommand{\shreyas}[1]{{\bf (#1)}}
\newcommand{\shreyas}[1]{}
\definecolor{tableShade}{gray}{0.8}

@@

%The spontaneous waving of marine grass is thought to be due to a Kelvin-Helmholtz instability resulting from an inflection point near the canopy top in the flow profile.
%We study the mechanism behind \monami, a phenomenon associated with the synchronous waving of marine grass. Linear stability analysis of a mathematical model incorporating grass 
%drag as a continuum field has been 
%carried out. Our analysis predicts a critical threshold condition for flow instability leading to \monami and associated waving frequency
%which compares well with the experimental observations. Further analysis predicts the existence of two distinct modes of instability which we refer as ``Mode 1'' and ``Mode 2''.
%Mode 1 has many characteristics similar to the Kelvin-Helmholtz instability where shear layer formed near grass tip due to grass drag discontinuity plays a critical role. 
%Mode 2 can be interpreted as instability of 
%inviscid flow through a grass bed where shear layer doesn't play an important role. 
%The occurrence of these modes for flow instability depends on submergence ratio and the grass 
%number density. Asymptotic analysis along with numerical computation in the limit of dense grass bed has been carried out. The analysis reveals that Mode 1 represents flow instability on the 
%scale of shear layer which is formed near grass tip due to drag discontinuity near the canopy top, whereas Mode 2 can be understood as instability on the scale of water column. 
%For the parameter regime typical of marine grass beds, the two modes merge and are indistinguishable.

@@

We show that the threshold flow condition for the spontaneous waving of marine grass arises due to the drag exerted by the vegetation on the flow. We show this by developing a self-consistent model of the flow past rigid vegetation. Linear stability analysis of the steady flow resulting from this model naturally leads to the threshold condition, and as well as the origin of the mechanistic threshold. Asymptotic analysis of the model for dense grass reveals two unstable modes. One of these modes shares many common features with the Kelvin-Helmholtz instability, but is demonstrated to be different from it. The predictions of the model for threshold condition and the resulting dominant frequency of waving agree with laboratory observations. 
\end{abstract}
\maketitle
%\section{Introduction}
Sea grasses exhibit rich set of dynamics due to their interaction with the flow of water and can considerably affect surrounding hydrodynamic conditions.
Resulting changes in hydrodynamic conditions can influence number of environmental variables and processes such as transport of sediments, contaminants, dissolved oxygen, plant growth, and biomass production, etc \cite{Fonseca87,Nepf99,Nepf2012}. 
Flexible grass canopies respond to steady currents in the form of coherent large amplitude oscillations, known as \monami \cite{AckermanOkubo93}. Such motion is considered to play a central role in the recruitment 
of microscopic marine organisms, such as blue mussel larvae\cite{Grizzle96}.
The hydrodynamic mechanism underlying \monami is the focus of this paper.

Existing explanation of \monami invokes the existence of strong shear near the top of the canopy \cite{Ghisal02,Raupach96} due to different amounts of drag experienced by fluid in and above the canopy. 
This shear layer is assumed to become unstable to coherent vortices through a mechanism similar to Kelvin-Helmholtz instability. 
Influence of these coherent eddies over sea grasses is manifested in their large amplitude synchronous oscillations.

@@

While the shear layer model successfully appears to predict frequency of \monami for number of experimental observations, several aspect of existing theory remain unsatisfactory. 
First, while the free shear layer is established due to the vegetation drag, the linear stability analysis\cite{Raupach96} does not take this drag into account, rendering this explanation inconsistent. 
Second, classical free shear flow is known to be unstable for all Reynolds number\cite{drazin}, whereas observations in the lab\cite{Ghisal02} and the field\cite{Grizzle96} indicate the existence of a threshold 
flow speed below which waving is not observed. 
Finally, the thickness of the free shear layer, defined as the momentum thickness of the boundary layer near the grass tip, is in many cases comparable to the unvegetated layer thickness,
and therefore inconsistent with the assumptions of the free shear layer instability. 
% These drawbacks of existing theory suggest that flow through vegetation requires further investigation for better understanding of the phenomenon.

In this paper we present a mathematical model exhibiting a rich behavior, which is not only consistent within itself, but also with lab experiments and field observations of the phenomenon.
The synchronous waving may be understood to be arising from the linear hydrodynamic instability of a uniform flow over the grass bed, such that the principal role of the vegetation is to exert a drag on the flow. 
A threshold flow speed arises naturally for the onset of the instability; this threshold condition and the resulting waving frequency compares well with experimental observations.
The vegetation drag introduces a damping mechanism in the flow, in addition to the viscous damping, and overcoming this damping is a primary consideration that sets the threshold flow.
We deduce this numerically and through an asymptotic analysis of the dense grass limit.

The drag exerted by the grass bed on the flow is central to the hypothesized mechanism\cite{Ghisal02} of the instability underlying \monami because it realizes the shear layer structure of the flow. 
The flexibility of the grass blades, on the other hand, is considered to merely aid in visualizing the flow structure resulting from the instability\cite{Nepf2012}. 
It has been demonstrated that the instability and the resulting flow structures are present even when the grass blades are replaced with rigid dowels\cite{Ghisal02}. 
Therefore, in the interest of deriving the simplest mathematical model for the linear instability, we consider the grass bed to form a rigid structure that exchanges momentum with the surrounding fluid but does not deform. 
Indeed, a more detailed calculation undertaken for terrestrial grass demonstrated\cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, 
the flow instability is similar to the instability of a shear layer.

\begin{figure}
\begin{center}
\includegraphics[]{Critical_Re_vs_delta_noshear} \\
\vspace{-6mm} \hspace{-3mm}
\includegraphics[]{K_vs_shear_width_noshear}
\end{center}
\caption{Critical Reynolds number with flow (top) and the corresponding marginally stable wave number (bottom) for different submergence ratio as a function of vegetation density
 parametrized by the boundary layer thickness. Parameters estimated from experiments reported by Ghisalberti and Nepf\cite{Ghisal02} to exhibit or suppress synchronous waving are 
also included in the top panel. Inset compares experimental observations 
of the dominant frequency $f_o$ (in Hz) measured with the predictions from the solution of \eqref{Orr-somerfield}. The experimental data in the inset in obtained from publications by 
Ghisalberti and Nepf \cite{Ghisal02} and Vivoni \cite{Vivoni98}. In order to estimate the $\Rey$ for these experiments, a representative value of $\mu=0.1$ Pa~s was assumed.}
\label{Re_vs_delta}
\end{figure}
We use a mean field model for the coupling between the flow and the canopy. 
Drag force exerted by the grass is approximated via a continuous body force $\mathbf{f}=-N_g\mathbf{f_d}$ in the fluid momentum balance\cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06} as
\begin{equation}
\rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad P+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}, \quad \grad\cdot\bu = 0,
\label{navier-stokes}
\end{equation}
where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, $N_g$ the grass number density per unit area, $\rho$ the fluid density, $\mathbf{u}$ the velocity, 
$P$ the pressure, $\mu$ the turbulent viscosity and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow on the scale of a grass blade is expected to be O($10^2$-$10^3$), and hence the drag force is dominated by form drag, and is modeled as 
$\mathbf{f_{d}}=C_N \rho u_{N}^{2}d\bn + C_{T}\rho u_{T}^{2}d\bt$ \shreyas{citation needed} 
where $d$ is average width of grass blade $C_{N}$ and $C_{T}$ are normal and tangential drag coefficients respectively; $u_{T}$, $u_{N}$ are velocity components along and
normal to grass while $\bt,\bn$ being unit vector along and normal to grass. 
We expect $C_T \ll C_N$ and take $C_T=0$ for rest of analysis. 
In the field, both $C_N$ and $N_g$ vary with distance from the bottom but we do not expect these variation to be central to the mechanism and therefore take $C_N$ and $N_g$ to be constants. 
Based on previous experiments\cite{Vivoni98}, we take $C_N \approx 1$.

For consistency of analysis, we calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies $-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$, where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$.
The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{basicflow}. 
Flow consist of a region within grass of approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from balance of drag force with pressure gradient and a parabolic velocity profile in unvegetated region due to balance of viscous force and pressure gradient. 
Continuity of flow speed and shear stress at canopy top gives rise to a boundary layer in the flow profile near the tip of the grass.

The boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and may therefore be identified as a free shear layer\cite{Ghisal02} in the previous explanation of \monami. 
In our model, this leads to a shear layer thickness $\delta = (\Rey\Ndg)^{-1/3} H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, and $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity. 
This dependence of the boundary layer thickness on the grass density $N_g$, shown in Figure \ref{basicflow}, gives us a way to systematically control it and investigate its effect on the instability mechanism.
The asymptotic regime of a thin boundary layer is expected to hold for $\Rey \Ndg \gtrsim 100$; departures from this scaling can be seen in the inset of figure \ref{basicflow}.

@@

%<<<<<<< HEAD
% A threshold in $\Rey$ for the onset of the instability naturally emerges from the solution of \eqref{Orr-somerfield}. 
% The primary reason for the existence of critical $\Rey$ is the presence of dissipation of flow instability 
% due to grass drag. Indeed a calculation undertaken for terrestrial canopy shows the suppression of growth rate of instability
% due to grass drag\cite{Delangre06}. Absence of threshold in the previous analysis can be attributed to the assumptions such 
% as approximationn of mean flow with the free shear layer profile and the exclusion of grass drag in the stability analysis 
% \cite{Raupach96,Delangre06}.
% The dependence of this critical $\Rey$, and the corresponding fastest growing wavenumber $k$, 
% on the $\delta/H$ and $\hg/H$ is shown in figure \ref{Re_vs_delta}, and compared with experimental observation of the 
% threshold\cite{Ghisal02}. The good agreement between predicted and estimated threshold suggest that the grass drag plays a 
% crucial role in setting up the flow instability.
% 
% Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, is also shown in 
% Figure \ref{Re_vs_delta}. The observed frequencies are associated with the peaks in velocity spectra, frequency of monami and frequency of vortex passage \cite{Ghisal02}\shreyas{cite someone}. The deviation of predicted frequency from the observed may be attributed to the various 
% simplification we have made in our model. In a real canopy, the drag coefficients are known to
% vary from bottom to tip of the grass blades\cite{Vivoni98,Nepf00}. The turbulent viscosity is also known to vary between the flow of unvegetated region and flow through the canopy\cite{Ghisal02}. Although these variation
% are not central to the mechanism leading to the instability of flow, we believe an improved analysis of flow with these variation in the model might lead to a better agreement between the
% observed frequency and the predicted frequency.
% =======
Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in
the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, 
is also shown in Figure \ref{Re_vs_delta}. 
The observed frequencies are associated with the peaks in velocity spectra, frequency of \monami and frequency of 
vortex passage \cite{Ghisal02}, and compare well with the predicted frequencies.

A threshold in $\Rey$ of O(100), above which the flow is unstable (Re($\sigma$)$>$0), naturally emerges from the 
solution of \eqref{Orr-somerfield}. 
The dependence of this critical $\Rey$, and the corresponding fastest growing wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown 
in figure \ref{Re_vs_delta}, and is found to compare well with experimental observations\cite{Ghisal02}.
The threshold Reynolds number increases with the grass density, indicating a competition between the 
destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
Previous linear stability calculations undertaken for terrestrial grass do not exhibit a threshold flow condition 
either due to the exclusion of the vegetation drag in their model\cite{Raupach96}, or because of the over-simplifying 
assumptions such as a tanh- or piece-wise linear mean velocity profile\cite{Raupach96,Delangre06}.
Regardless of the assumptions, the introduction of vegetation drag damped the linear modes.
%>>>>>>> bd11ad10d1dc4414f38f11c1e91e5e1d200223a2

\begin{table*}
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.4}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{$-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left[({\rho U_0 H}/\mu) (C_N d N_g H)\right]^{-1/3}$} \\
 Linearized dynamics	& Inviscid Rayleigh's equation			& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Critical parameters	& none						& $\Rey$, $\delta/H$ 	& $\Rey \delta^{3/2}$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto \delta^{-1}$		& $\propto \delta^{-1}$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& ~~~~yes, near $y=\hg$~~~~			& no
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}

To better understand the mechanism of waving and the origin of the threshold, we consider the behavior of the instability as a function of the vegetation drag.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical grass density discontinuously jumps and remains $O(1)$. 
To aid in explaining the behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in figure \ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two. We refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the grass density increases, causing the discontinuity in the most unstable mode.

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 2 has a threshold $\Rey \sim ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in figure \ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of \eqref{Orr-somerfield}.
Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the grass is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
Using this estimate, and $\Rey\gg 1$, simplifies \eqref{Orr-somerfield} within the grass to $\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi$, while in the unvegetated region it simplifies to the Rayleigh equation $ \left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$. 
This analysis predicts that the mode shape would converge as $\Rey\gg 1$ and $\Ndg \gg 1$ but $\Rey/\Ndg \sim O(1)$, in agreement with our numerical results shown in figure \ref{Asymptotic_mode}.

@@

\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness 
for representative values of grass density. Mode 1 is shown in solid and Mode 2 is shown in dashed. 
The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specifying in terms of $\delta/H$) 
but $\Rey/\Ndg = O(1)$. The mode shapes approach each other for these small values of $\delta/H$ indicating that 
the large grass density asymptote is reached. Mode 2 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. Inset shows $|\phi|$ for 
Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit 
has been reached. The limit is not yet reached for the case $\delta/H = 0.05$ due to the influence of bottom boundary; 
the grass height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}

>>>>

\documentclass[aps,prl,reprint,twocolumn,showpacs,superscriptaddress,10pt]{revtex4-1}  % for review and submission
%\documentclass[aps,twocolumn,floatfix,prl,10pt]{revtex4-1}
%\documentclass[letterpaper,10pt,prl,twocolumn,aps]{revtex4-1}
%\documentclass[aps,preprint,floatfix,prl]{revtex4-1}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{color}
\usepackage{longtable}
\usepackage{array}
\usepackage{amssymb}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

\graphicspath{ {images/}{images/Growth_Rate_Cont_Plots/} }
\newcommand{\bn}{{\boldsymbol{\hat{n}}}}
\newcommand{\bt}{{\boldsymbol{\hat{t}}}}
\newcommand{\bu}{\mathbf{u}}
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand{\del}{\partial}
\newcommand{\hg}{h_g}
\newcommand{\Rey}{\text{R}}
\newcommand{\Ndg}{\tilde{N}_g}
% \newcommand{\shreyas}[1]{{\bf (#1)}}
\newcommand{\shreyas}[1]{}
\definecolor{tableShade}{gray}{0.8}

@@

%The spontaneous waving of marine grass is thought to be due to a Kelvin-Helmholtz instability resulting from an inflection point near the canopy top in the flow profile.
We study the mechanism behind Mo-nami, a phenomenon associated with the synchronous waving of marine grass. Linear stability analysis of a mathematical model incorporating grass 
drag as a continuum field has been 
carried out. Our analysis predicts a critical threshold condition for flow instability leading to Mo-nami and associated waving frequency
which compares well with the experimental observations. Further analysis predicts the existence of two distinct modes of instability which we refer as ``Mode 1'' and ``Mode 2''.
%Mode 1 has many characteristics similar to the Kelvin-Helmholtz instability where shear layer formed near grass tip due to grass drag discontinuity plays a critical role. 
%Mode 2 can be interpreted as instability of 
%inviscid flow through a grass bed where shear layer doesn't play an important role. 
The occurrence of these modes for flow instability depends on submergence ratio and the grass 
number density. Asymptotic analysis along with numerical computation in the limit of dense grass bed has been carried out. The analysis reveals that Mode 1 represents flow instability on the 
scale of shear layer which is formed near grass tip due to drag discontinuity near the canopy top, whereas Mode 2 can be understood as instability on the scale of water column. 
For the parameter regime typical of marine grass beds, the two modes merge and are indistinguishable.

@@

\end{abstract}
\maketitle
%\section{Introduction}
Sea grasses exhibit rich set of dynamics due to their interaction with the flow of water and can considerably affect surrounding hydrodynamic conditions.
Resulting changes in hydrodynamic conditions can influence number of environmental variables and processes such as transport of sediments, contaminants, dissolved oxygen, plant growth, and biomass production, etc \cite{Fonseca87,Nepf99,Nepf2012}. 
Flexible grass canopies respond to steady currents in the form of coherent large amplitude oscillations, known as mo-nami \cite{AckermanOkubo93}. Such motion is considered to play a central role in the recruitment 
of microscopic marine organisms, such as blue mussel larvae\cite{Grizzle96}.
The hydrodynamic mechanism underlying monami is the focus of this paper.

Existing explanation of monami invokes the existence of strong shear near the top of the canopy \cite{Ghisal02,Raupach96} due to different amounts of drag experienced by fluid in and above the canopy. 
This shear layer is assumed to become unstable to coherent vortices through a mechanism similar to Kelvin-Helmholtz instability. 
Influence of these coherent eddies over sea grasses is manifested in their large amplitude synchronous oscillations.

@@

While the shear layer model successfully appears to predict frequency of monami for number of experimental observations, several aspect of existing theory remain unsatisfactory. 
First, while the free shear layer is established due to the vegetation drag, the linear stability analysis\cite{Raupach96} does not take this drag into account, rendering this explanation inconsistent. 
Second, classical free shear flow is known to be unstable for all Reynolds number\cite{drazin}, whereas observations in the lab\cite{Ghisal02} and the field\cite{Grizzle96} indicate the existence of a threshold 
flow speed below which waving is not observed. 
Finally, the thickness of the free shear layer, defined as the momentum thickness of the boundary layer near the grass tip, is in many cases comparable to the unvegetated layer thickness,
and therefore inconsistent with the assumptions of the free shear layer instability. 
These drawbacks of existing theory suggest that flow through vegetation requires further investigation for better understanding of the phenomenon.

In this paper we present a mathematical model exhibiting a rich behavior, which is not only consistent within itself, but also with lab experiments and field observations of the phenomenon.
The synchronous waving may be understood to be arising from the linear instability of a uniform flow over the grass bed. 
The threshold condition for the onset of the instability and the resulting waving frequency compares well with experimental observations.
In the limit of a dense grass bed, two distinct unstable modes can be asymptotically distinguished, with one of them sharing many characteristics with the Kelvin-Helmholtz instability. 
In the parameter regime typical of marine grass beds, the two modes merge and are indistinguishable.

The drag exerted by the grass bed on the flow is central to the hypothesized mechanism\cite{Ghisal02} of the instability underlying monami because it realizes the shear layer structure of the flow. 
The flexibility of the grass blades, on the other hand, is considered to merely aid in visualizing the flow structure resulting from the instability\cite{Nepf2012}. 
It has been demonstrated that the instability and the resulting flow structures are present even when the grass blades are replaced with rigid dowels\cite{Ghisal02}. 
Therefore, in the interest of deriving the simplest mathematical model for the linear instability, we consider the grass bed to form a rigid structure that exchanges momentum with the surrounding fluid but does not deform. 
Indeed, a more detailed calculation undertaken for terrestrial grass demonstrated\cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, 
the flow instability is similar to the instability of a shear layer.

\begin{figure}
\begin{center}
\includegraphics[]{Critical_Re_vs_delta_noshear} \\
\vspace{-6mm} \hspace{-3mm}
\includegraphics[]{K_vs_shear_width_noshear}
\end{center}
\caption{Critical Reynolds number with flow (top) and the corresponding marginally stable wave number (bottom) for different submergence ratio as a function of vegetation density
 parametrized by the boundary layer thickness. Parameters estimated from experiments reported by Ghisalberti and Nepf\cite{Ghisal02} to exhibit or suppress synchronous waving are 
also included in the top panel. In order to estimate the $\Rey$ for these experiments, a representative value of $\mu=0.1$ Pa~s was assumed. Inset compares experimental observations 
of the dominant frequency $f_o$ (in Hz) measured with the predictions from the solution of \eqref{Orr-somerfield}. The experimental data in the inset in obtained from the articles by 
Ghisalberti \cite{Ghisal02} and Vivoni \cite{Vivoni98}.}
\label{Re_vs_delta}
\end{figure}
We use a mean field model for the coupling between the flow and the canopy. 
Drag force exerted by the grass is approximated via a continuous body force $\mathbf{f}=-N_g\mathbf{f_d}$ in the fluid momentum balance\cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06} as
\begin{equation}
\rho \left(\bu_{t}+\bu.\grad\bu \right) = -\grad P+\mu\grad^{2}\bu +\mathbf{f}+\rho\mathbf{g}, \quad \grad\cdot\bu = 0,
\label{navier-stokes}
\end{equation}
where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, $N_g$ the grass number density per unit area, $\rho$ the fluid density, $\mathbf{u}$ the velocity, 
$P$ the pressure, $\mu$ the turbulent viscosity and $\mathbf{g}$ the acceleration due to gravity. 
The Reynolds number of the flow on the scale of a grass blade is expected to be O($10^2$-$10^3$), and hence the drag force is dominated by form drag, and is modeled as 
$\mathbf{f_{d}}=C_N \rho u_{N}^{2}d\bn + C_{T}\rho u_{T}^{2}d\bt$ \shreyas{citation needed} 
where $d$ is average width of grass blade $C_{N}$ and $C_{T}$ are normal and tangential drag coefficients respectively; $\bu_{T}$, $\bu_{N}$ are velocity vector along and
normal to grass while $\bt,\bn$ being unit vector along and normal to grass. 
We expect $C_T \ll C_N$ and take $C_T=0$ for rest of analysis. 
In the field, both $C_N$ and $N_g$ vary with distance from the bottom but we do not expect these variation to be central to the mechanism and therefore take $C_N$ and $N_g$ to be constants. 
Based on previous experiments\cite{Vivoni98}, we take $C_N \approx 1$.

For consistency of analysis, we calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies $-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$, where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$.
The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{basicflow}. 
Flow consist of a region within grass of approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from balance of drag force with pressure gradient and a parabolic velocity profile in unvegetated region due to balance of viscous force and pressure gradient. 
Continuity of flow speed and shear stress at canopy top give rise to a boundary layer in the flow profile near the tip of the grass. 
The boundary layer near the canopy top results from purely local dynamics independent of the influence of the boundaries, and may therefore be identified as a free shear layer\cite{Ghisal02} in the previous explanation of monami. 
In our model, this leads to a shear layer thickness $\delta = (\Rey\Ndg)^{-1/3} H$, where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, and $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity. 
This dependence of the boundary layer thickness on the grass density $N_g$, shown in Figure \ref{basicflow}, gives us a way to systematically control it and investigate its effect on the instability mechanism.

@@

A threshold in $\Rey$ for the onset of the instability naturally emerges from the solution of \eqref{Orr-somerfield}, which were not predicted in the previous calculation of 
flow instability of free shear layer profile \cite{Raupach96}\cite{Delangre06}.
The dependence of this critical $\Rey$ and the corresponding fastest growing wavenumber $k$, on the $\delta/H$ and $\hg/H$ is shown in figure \ref{Re_vs_delta}, 
and compared with experimental observation of the threshold\cite{Ghisal02}. 
The primary reason for existence of this threshold is the dissipation due to grass drag which has to be overcome by flow instability.
The increase in the value of critical $Rey$ with the increase in grass number density can be attributed to the existence of increased dissipation  Dissipation due to grass drag increases with the increase in $\tilde{Ng}$; hence a higher $\Rey$ for flow instability to occur.

Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in the lab scale experiments, for cases where the 
vegetation was sufficiently dense to be modeled by a continuum drag field, is also shown in 
Figure \ref{Re_vs_delta}. The observed frequencies are associated with the peaks in velocity spectra, frequency of monami and frequency of 
vortex passage \cite{Ghisal02}\shreyas{cite someone}. The deviation of predicted frequency from the observed may be attributed to the various 
simplification we have made in our model. In a real canopy, the drag coefficients are known to
vary from bottom to tip of the grass blades\cite{Vivoni98,Nepf00}. The turbulent viscosity is also known to vary between the flow of unvegetated region and 
flow through the canopy\cite{Ghisal02}. Although these variation
are not central to the mechanism leading to the instability of flow, we believe an improved analysis of flow with these variation in the model might lead to 
a better agreement between the
observed frequency and the predicted frequency.

To better understand the mechanism of waving, we consider the behavior of the instability, especially the fastest growing wavenumber, as a function of the vegetation drag, especially as it becomes 
asymptotically large, i.e. $\Ndg \gg 1$ (or $\delta \ll H$).
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical grass density discontinuously jumps and remains $O(1)$. 
To aid in explaining the behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$ \shreyas{Make sure to remone $H$ from $kH$ because $k$ is dimensionless.}, 
for different $\hg/H$ and $\Ndg$ in figure \ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two. We refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the grass density increases, causing the discontinuity in the most unstable mode.

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 2 has a threshold $\Rey \sim ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in figure \ref{Re_vs_delta}, which can be understood by asymptotic analysis of \eqref{Orr-somerfield}.
Noting that for large $\Ndg$, the non-dimensional steady flow velocity inside the grass is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
Using this estimate, and $\Rey\gg1$, \eqref{Orr-somerfield} within the grass simplifies into $\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi$ \shreyas{check sign}, while in the unvegetated region it simplifies to 
the Rayleigh equation $ \left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$. 
This analysis predicts that the mode shape would converge as $\Rey\gg 1$ and $\Ndg \gg 1$ but $\Rey/\Ndg \sim O(1)$, in agreement with out numerical results shown in figure \ref{Asymptotic_mode}.

@@

\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of grass density. 
Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specifying in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The mode shapes approach each other for these small values of $\delta/H$ indicating that the large grass density asymptote is reached. Mode 2 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. 
The limit is not yet reached for the case $\delta/H = 0.01$ due to the influence of bottom boundary; the grass height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}
